PeakNet: An Autonomous Bragg Peak Finder with Deep Neural Networks

### 2023-06-28 manuscript.tex
Correct format for arxiv submission.

Serial crystallography at X-ray free electron laser (XFEL) sources has experienced tremendous progresses in achieving high data rate in recent times, offering potential to enable novel scientific investigations, such as imaging faster molecular events.  However, this approach also generates a significant amount volume of data with sparsely distributed scientific signals, posing challenges in data reduction and real-time feedback.  To address these concerns, a rapid and efficient Bragg peak finding process serves as the pivotal initial step, effectively addressing both data storage concerns and real-time feedback requirements simultaneously.  Here, we present \peaknet{}, a autonomous Bragg peak finder that utilizes deep neural networks, eliminating the need for manual parameter tuning common in conventional peak finders and the laborious task of manually masking pixels affected by artifact scattering, such as nozzle scattering rings.  \peaknet{} also exhibits exceptional runtime efficiency, processing a $1920 \times 1920$ image around 90 ms on an NVIDIA 1080 Ti GPU, with the potential for further enhancements through parallelized analysis or GPU stream processing.  With its ability to scale linearly with respect to data volume, \peaknet{} is well-suited for real-time serial crystallography data analysis at high data rates.

@@

Serial crystallography with X-ray free electron lasers (XFEL) has enabled the structural determination of three-dimensional macromolecules from radiation damage-free samples at room temperature, an approach commonly known as \textit{diffraction-before-destruction} \citep{neutzePotentialBiomolecularImaging2000, chapmanFemtosecondDiffractiveImaging2006,chapmanFemtosecondXrayProtein2011}.  Time-resolved serial crystallography at XFEL facilities, like the Linac Coherent Light Source (LCLS), allows the investigation of biochemical reactions with femtosecond-scale time resolutions.  This technique is typically referred to as serial femtosecond crystallography (SFX), which has led to many significant structural studies \citep{kupitzSerialTimeresolvedCrystallography2014, nangoThreedimensionalMovieStructural2016,pandeFemtosecondStructuralDynamics2016a, youngStructurePhotosystemII2016,sugaLightinducedStructuralChanges2017, kernStructuresIntermediatesKok2018,ibrahimUntanglingSequenceEvents2020, sugaTimeresolvedStudiesMetalloproteins2020} since the first published work at LCLS \citep{aquilaTimeresolvedProteinNanocrystallography2012}.  In the meantime, the data rate at XFEL facilities have also gradually increased from hundreds of Hz to the MHz range, with the European X-ray Free Electron Laser (EuXFEL) being the first MHz facility available to users and LCLS-II set to launch in 2023.

At the MHz data rate, novel real-time data analysis approaches are needed to tackle two major challenges that arise in SFX experiments, namely data reduction and real time feedback.  One common approach for finding hits in serial crystallography data is to utilize existing Bragg peak finders like the Cheetah peak finder \citep{bartyCheetahSoftwareHighthroughput2014}, the Robust Peak Finder (RPF) \citep{hadian-jaziPeakfindingAlgorithmBased2017}, or the \psocake{} \citep{yoonPsocakeGUIMaking2020} peak finder.  Additionally, a deep neural network-based hit finder for serial crystallography has been reported to achieve a processing rate of 1.3 kHz by analyzing downsized 180 $\times$ 180 images using a single GPU \citep{keConvolutionalNeuralNetworkbased2018}.  Secondly, real-time feedback in serial crystallography experiments plays a crucial role by providing users with important information about the experimental conditions.  This enables them to make timely decisions that optimize the data acquisition process. For instance, it is essential to fine-tune the hit rate to maximize the utilization of samples. Additionally, it is important to monitor factors such as the current diffraction limit, crystalline mosaicity, reciprocal space coverage, and other relevant parameters over time. Therefore, relying solely on hit finding is insufficient to gain the necessary insights, making real-time Bragg peak finding vital for delivering prompt and comprehensive feedback. However, the peak finder mentioned earlier in \psocake{} operates at a rate of approximately 5 Hz on a data center CPU, which could lead to significant costs when scaling up the process.

In this work, we introduce \peaknet{}, a solution that effectively addresses the challenge of autonomously and efficiently identifying Bragg peaks in high data rate scenarios.  Our approach involves transforming the peak finding problem into semantic segmentation tasks, where different regions within diffraction images are classified and segmented using deep learning networks.  By leveraging this process, \peaknet{} autonomously detects Bragg peaks while effectively identifying strong artifact scattering, eliminating the need for human intervention.  Subsequently, connected areas of pixels are examined and treated as individual peaks, a task easily accomplished through connected component analysis \citep{weaverCentrosymmetricCrossSymmetricMatrices1985}.  The coordinates of the peaks are determined by calculating the center of mass of all pixels belonging to each identified peak.  Notably, the entire peak finding process is performed on GPUs, offering a cost-effective solution for achieving scalability.

@@

While template matching and region-growing techniques require users to provide priori information like threshold values, Robust Peak Finder (RPF) \citep{hadian-jaziPeakfindingAlgorithmBased2017, hadian-jaziDataReductionSerial2021} has demonstrated proficient peak finding capabilities with minimal user inputs.  This technique utilizes the Modified Selective Statistical Estimator (MSSE) method to segment between background pixels and actual peaks.  Furthermore, the RPF method can also parallelize the processing of diffraction patterns, making it ideal for real-time data analysis.

\psocake{} is a software specifically designed to streamline high throughput data reduction and analysis of XFEL experiments, and is routinely used at SLAC National Accelerator laboratory (SLAC) and Pohang Accelerator Laboratory (PAL).  The built-in peak-finding algorithm \citep{shinDataAnalysisUsing2018} includes calibrating raw analog digital units (ADUs); applying optional background correction and bad pixel masks; identifying possible peaks; calculating their SNR; determining Bragg spot sizes; and selecting peaks based on size, total intensity and SNR.

@@

The current peak finding process of \peaknet{} comprises two steps, as depicted in Fig. \ref{fig : peak finding}.  Firstly, a segmentation map is obtained from the input image using deep neural networks.  Subsequently, peak positions are determined based on the generated segmentation map using connected component analysis \citep{weaverCentrosymmetricCrossSymmetricMatrices1985}.  Importantly, \peaknet{} has the capability to process inputs with arbitrary image sizes in a batch manner, enabling efficient analysis.

In this section, our focus is to provide a comprehensive understanding of the underlying deep neural network components in \peaknet{}.  We will specifically discuss the network architecture, data augmentation techniques, and the loss function employed during the training process.  Furthermore, we will emphasize the use of iterative data curation in addressing the outlier problem.

\begin{figure*}[!ht]
\includegraphics[width=\textwidth,keepaspectratio]
{./figures/peaknet_steps.pdf}
\caption{The peak finding process of \peaknet{} consists of two steps: (1)
obtaining a segmentation map from the input using the underlying deep neural
networks; (2) identifying peak positions based on the generated segmentation map
using connected component analysis.  It is worth noting that \peaknet{} is
capable of accommodating input with arbitrary image sizes and batch sizes.  }
\label{fig : peak finding}
\end{figure*}

@@

It is essential to acknowledge that our neural network architecture is primarily based on the original ``Attention U-Net" design \citep{oktayAttentionUNetLearning2018}.  However, we want to point out that the three-component design, consisting of the ``backbone", ``neck" and ``head", offers the flexibility to replace these components with different architectures.  For instance, the ``backbone" can be substituted with RegNet \citep{radosavovicDesigningNetworkDesign2020}, while the ``neck" can be replaced with other multi-resolution feature fusion methods like feature pyramid networks (FPN) \citep{linFeaturePyramidNetworks2017} or bidirectional feature pyramid networks (BiFPN) \citep{tanEfficientDetScalableEfficient2020}.  Additionally, the ``head" can be extended to perform additional tasks, if desired, such as predicting the likelihood of the input image being indexable.  This modular approach provides flexibility and allows for customization based on specific requirements.

\begin{figure*}[!ht]
\includegraphics[width=\textwidth,keepaspectratio]
{./figures/network_arc.pdf}
\caption{Neural network architecture: Attention U-Net with residual connections.
(a) Overall end-to-end network architecture.  (b) The residual double
convolutional block.  (c) The gated feature fusion block for merging
multi-resolution features.}
\label{fig : network arc}
\end{figure*}

@@

One main technical issue in training neural networks for Bragg peak segmentation is the extreme peak-background class imbalance (e.g., $> 10^3 : 1$), resulting in reduced prediction accuracy.  This may not pose as a problem in previous peak finders like BraggNet, which operates on a much smaller area of $32 \times 32$, whereas \peaknet{} is trained on larger detectors, such as Rayonix MX340-XFEL detector with a dimension of 1920 $\times$ 1920, even after downsizing by a factor of 4.  To mitigate this challenge, focal loss was introduced to address the extreme class imbalance problem for object detectors \citep{linFocalLossDense2018}.  More specifically, we employed a categorical focal loss.

\begin{align}
\text{CFL} &= - \sum_{i = 1}^{N}\sum_{j = 1}^{C} 
            \alpha_j \cdot y_j^{(i)} \cdot (1-\hat{p}_j^{(i)})^\gamma 
            \cdot \log{\hat{p}_j^{(i)}} \\
\hat{p}_j^{(i)} &= f_\theta(x_j^{(i)})
\end{align}

where CFL stands for categorical focal loss, $\alpha_j$ is a balancing factor for each class, $y_j^{(i)}$ is the grouth truth label of the $j$-th class for the $i$-th example.  $p_j^{(i)}$ is the predicted probability of the $j$-th class for the $i$-th example, calculated by the neural network $f$ parameterized by $\theta$ given input $x_j^{(i)}$.  $\gamma$ is a parameter controlling the extent to which the weight of well-classified examples, such as background pixels, is reduced.  In our neural network training, we chose $\gamma = 2.0$, essentially, this means $(1-\hat{p} _j^{(i)}) ^\gamma$ is solely responsible for down-weighting easy examples, as $\hat{p} _j^{(i)}$ is usually large in this case.  Finding the ideal values for $\alpha$ can be a delicate process with only marginal returns, as highlighted in the original focal loss study \citep{linFocalLossDense2018}.  In order to fully utilize focal loss, we decided to use a fixed trivial value of $\alpha = 1.2$ for all examples, primarily for the sake of completeness.  This choice essentially treats $\alpha$ as a simple scaling factor without significant impact.  It is worth noting that in most cases, the sum of all $\alpha_j$ values, denoted as $\sum_{j = 1}^{C} \alpha_j$, is expected to equal 1.

@@

We applied three data augmentation techniques to each diffraction pattern in our dataset, including random in-plane rotation, random shifting in both horizontal and vertical directions and random masking.  Random in-plane rotation and shifting augment the training data, reducing the model's tendency to memorize features at specific locations.  This variability promotes more generalized feature extraction and improved generalization performance.  Random masking involves randomly obscuring parts of the input data, forcing the model to learn other relevant features that enhance its predictive capabilities during training.  The result of applying data augmentation to an arbitrary example is shown in Fig. \ref{fig : data aug}.

\begin{figure*}[!ht]
\includegraphics[width=\textwidth,keepaspectratio]
{./figures/data_aug.pdf}
\caption{Data augmentation is utilized on an input (a) and its corresponding
labeled segmentation map (b). The zoomed-in views of (a) and (b) are represented
by (c) and (d), respectively. The segmentation label distinguishes red pixels as
peaks and blue pixels as artifact scattering, attributed to ``nozzle scattering"
in this particular scenario.}
\label{fig : data aug}
\end{figure*}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\columnwidth,keepaspectratio,trim={1in 0in 1in 0in}]
{./figures/data_engine.pdf}
\caption{The process of iterative dataset and model improvement.}
\label{fig : data engine}
\end{figure}

@@

The primary goal of \peaknet{} is to autonomously detect Bragg peaks in diffraction images, devoid of any user input, such as algorithmic parameter tuning or manual artifact masking.  We implemented a two-step approach for peak detection, employing deep neural networks to perform semantic segmentation, which was subsequently followed by connected component analysis to extract image coordinates of Bragg peaks from the generated segmentation maps.  In this section, we begin by introducing the datasets used for training, validating, and testing \peaknet{}. Next, we present the evaluation results that demonstrate the predictive performance of \peaknet{}. Finally, we showcase its auto-masking capability and runtime efficiency.

@@

The dataset ``mfx13016" exhibits a notable characteristic of prominent nozzle scattering rings accompanied by bright Bragg peak-like spots located on the rings.  This characteristic presents two primary challenges for crystallography data analysis: increased difficulty in crystal indexing and misleading integration results caused by artifact scattering.  The current best practice involves manual masking of these artifact scattering, enabling data analysis to proceed as if they are absent, but this process is prone to errors and time-consuming.  Additionally, the final outcomes often hinge on the expertise of the individuals conducting the data processing, resulting in biases, scalability limitations, lack of standardization, and the possibility of human errors.  Nevertheless, the inherent challenges presented by this dataset offer an exceptional opportunity to testify the autonomous peak finding capability of \peaknet{}.

@@

The peak finding performance of peak finders is evaluated through two significant indicators from a user's standpoint: the number of successfully indexed hits and the merging statistics.  For our evaluation, we utilized the CrystFEL software suite \citep{whiteCrystFELSoftwareSuite2012} to perform crystal indexing, integration, and merging tasks.  We use \psocake{}'s peak finder as a baseline in this evaluation.

@@

\peaknet{} and \psocake{}, when manually fine-tuned, exhibit comparable peak finding performance in terms of the number of hits and the number of indexed images, as shown in Table \ref{tb : index}.  However, it is critical to emphasize that while \peaknet{} operates entirely autonomously, \psocake{}'s performance is contingent upon manual parameter fine-tuning and manual artifact scattering masking.  For example, an earlier work \citep{suSerialCrystallographyUsing2021}, which processed essentially the same experimental data, reported significantly lower hit count and indexed image count.  Consequently, this highlights \peaknet{}'s advantage in user-independent data processing.

Moreover, while users might prioritize the absolute number of indexed images as a more straightforward metric, the indexing rate provides valuable insights into the method's efficiency.  Both metrics jointly offer a more comprehensive evaluation of the peak finding method. In the case of \peaknet{}, both the consistently high number of indexed images and the good indexing rate testify to its effective and reliable autonomous peak finding process, positioning it as a highly viable option for Bragg peak finding.

@@

\multicolumn{1}{l}{\multirow{3}{*}{\peaknet{}}}                                                        & \multicolumn{1}{l}{Number of hits}       & \multicolumn{1}{l}{1205}    & \multicolumn{1}{l}{3248}    & \multicolumn{1}{l}{5797}    & \multicolumn{1}{l}{3949}    & \multicolumn{1}{l}{1184}    & \multicolumn{1}{l}{1030}    & 1223    \\ \cline{2-9}
\multicolumn{1}{l}{}                                                                                   & \multicolumn{1}{l}{Number of indexed}    & \multicolumn{1}{l}{860}     & \multicolumn{1}{l}{2856}    & \multicolumn{1}{l}{3433}    & \multicolumn{1}{l}{2401}    & \multicolumn{1}{l}{919}     & \multicolumn{1}{l}{953}     & 1169    \\ \cline{2-9}
\multicolumn{1}{l}{}                                                                                   & \multicolumn{1}{l}{Indexing rate}        & \multicolumn{1}{l}{0.71}    & \multicolumn{1}{l}{0.88}    & \multicolumn{1}{l}{0.59}    & \multicolumn{1}{l}{0.61}    & \multicolumn{1}{l}{0.78}    & \multicolumn{1}{l}{0.93}    & 0.96    \\ \hline

@@

\multicolumn{1}{l}{\multirow{3}{*}{\begin{tabular}[c]{@{}l@{}}\psocake{}\\ (fine-tuned)\end{tabular}}} & \multicolumn{1}{l}{Number of hits}       & \multicolumn{1}{l}{1193}    & \multicolumn{1}{l}{3087}    & \multicolumn{1}{l}{5769}    & \multicolumn{1}{l}{3926}    & \multicolumn{1}{l}{917}     & \multicolumn{1}{l}{1070}    & 1212    \\ \cline{2-9}
\multicolumn{1}{l}{}                                                                                   & \multicolumn{1}{l}{Number of indexed}    & \multicolumn{1}{l}{849}     & \multicolumn{1}{l}{2657}    & \multicolumn{1}{l}{3375}    & \multicolumn{1}{l}{2364}    & \multicolumn{1}{l}{760}     & \multicolumn{1}{l}{1005}    & 1202    \\ \cline{2-9}
\multicolumn{1}{l}{}                                                                                   & \multicolumn{1}{l}{Indexing rate}        & \multicolumn{1}{l}{0.71}    & \multicolumn{1}{l}{0.86}    & \multicolumn{1}{l}{0.59}    & \multicolumn{1}{l}{0.6}     & \multicolumn{1}{l}{0.83}    & \multicolumn{1}{l}{0.94}    & 0.99    \\ \hline

@@

Merging statistics confirms \peaknet{}'s robust performance in terms of consistency and quality in data processing.  For the thaumatin (TH) sample, \peaknet{} demonstrates marginally superior performance in terms of R-split values, correlation coefficients, $I/\sigma(I)$ values, redundancy, and completeness, achieving these results autonomously.  Notably, the R-split value for the TH sample is slightly lower for \peaknet{} (0.2605 vs. 0.2732), suggesting slightly better internal consistency. Similarly, correlation coefficients for \peaknet{} are marginally higher (CC$_1/2$: 0.9434 vs.  0.9368; CC*: 0.9853 vs. 0.9835), indicating a minor advantage in the precision of measurements.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=\textwidth,keepaspectratio]
{./figures/automask.agbehnate.pdf}
\caption{Examples of identifying scattering rings in Ag Behenate images using
\peaknet{}.}
\label{fig : automask 2}
\end{figure*}

@@

Peak finding can pose significant computational demands, particularly in high data rate experiments, highlighting the critical importance of an efficient peak finding algorithm.  A comparative analysis of runtime performance demonstrates the greater efficiency of \peaknet{} compared to other commonly used peak finders like the one within \psocake{}.

>>>>

Serial crystallography at X-ray free electron laser (XFEL) and synchrotron facilities has experienced tremendous progress in recent times enabling novel scientific investigations into macromolecular structures and molecular processes.  However, these experiments generate a significant amount of data posing computational challenges in data reduction and real-time feedback. Bragg peak finding algorithm is used to identify useful images and also provide real-time feedback about hit-rate and resolution. Shot-to-shot intensity fluctuations and strong background scattering from buffer solution, injection nozzle and other shielding materials make this a time-consuming optimization problem.  Here, we present \peaknet{}, an autonomous Bragg peak finder that utilizes deep neural networks. The development of this system 1) eliminates the need for manual algorithm parameter tuning, 2) reduces false-positive peaks by adjusting to shot-to-shot variations in strong background scattering in real-time, 3) eliminates the laborious task of manually creating bad pixel masks and the need to store these masks per event since these can be regenerated on demand.  \peaknet{} also exhibits exceptional runtime efficiency, processing a $1920 \times 1920$ pixel image around 90 ms on an NVIDIA 1080 Ti GPU, with the potential for further enhancements through parallelized analysis or GPU stream processing.  \peaknet{} is well-suited for expert-level real-time serial crystallography data analysis at high data rates.

@@

Serial crystallography with X-ray free electron lasers (XFEL) has enabled radiation damage-free structural determination of macromolecules at room temperature, an approach commonly known as \textit{diffraction-before-destruction} \citep{neutzePotentialBiomolecularImaging2000, chapmanFemtosecondDiffractiveImaging2006,chapmanFemtosecondXrayProtein2011}.  Time-resolved serial femtosecond crystallography at XFEL facilities, like the Linac Coherent Light Source (LCLS), allows the investigation of biochemical reactions with femtosecond-scale time resolution which has led to many significant structural studies \citep{kupitzSerialTimeresolvedCrystallography2014, nangoThreedimensionalMovieStructural2016,pandeFemtosecondStructuralDynamics2016a, youngStructurePhotosystemII2016,sugaLightinducedStructuralChanges2017, kernStructuresIntermediatesKok2018,ibrahimUntanglingSequenceEvents2020, sugaTimeresolvedStudiesMetalloproteins2020} since the first published work at LCLS \citep{aquilaTimeresolvedProteinNanocrystallography2012}.  In the meantime, the data rate at XFEL facilities have also gradually increased from hundreds of Hz to the MHz range, with the European X-ray Free Electron Laser (EuXFEL) being the first MHz facility available to users and LCLS-II set to launch in 2023.

At the MHz data rate, novel real-time data analysis approaches are needed to tackle two major challenges that arise in SFX experiments, namely data reduction and real time feedback.  The standard approach for data reduction is vetoing events where the x-ray pulse did not ``hit'' the sample. Typically, hit finding in serial crystallography utilizes Bragg peak finding algorithms implemented in such programs as Cheetah  \citep{bartyCheetahSoftwareHighthroughput2014}, DIALS \citep{winterDIALSImplementationEvaluation2018}, and \psocake{} \citep{yoonPsocakeGUIMaking2020}.  Additionally, a deep neural network-based hit finder for serial crystallography has been reported to achieve a processing rate of 1.3 kHz by analyzing downsized 180 $\times$ 180 images on a single GPU \citep{keConvolutionalNeuralNetworkbased2018}.  Real-time feedback in serial crystallography experiments plays a crucial role by providing users with important information about the experimental conditions.  This enables them to make timely decisions that optimize the data collection process. For instance, it is essential to fine-tune the hit rate to maximize the utilization of samples. Additionally, it is important to monitor factors such as the current diffraction limit, crystalline mosaicity, reciprocal space coverage, and other relevant parameters over time. Therefore, relying solely on hit finding is insufficient to gain the necessary insights, making real-time Bragg peak finding vital for delivering prompt and comprehensive feedback.

In this work, we introduce \peaknet{}, a solution that effectively addresses the challenge of autonomously and efficiently identifying Bragg peaks in high data rate scenarios.  Our approach involves transforming the peak finding problem into a semantic segmentation task, where different regions within diffraction images are classified and segmented using a deep neural network. \peaknet{} autonomously detects Bragg peaks while effectively identifying shot-to-shot variations of strong non-Bragg scattering artifacts, eliminating the need for human intervention.  Subsequently, connected areas of pixels are examined and treated as individual peaks, a task easily accomplished through connected component analysis \citep{weaverCentrosymmetricCrossSymmetricMatrices1985}.  The coordinates of the peaks are determined by calculating the center of mass of all pixels belonging to each identified peak.  Notably, the entire peak finding process is performed on GPUs, offering a cost-effective solution for achieving low latency.

@@

While template matching and region-growing techniques require users to provide prior information like threshold values, Robust Peak Finder (RPF) \citep{hadian-jaziPeakfindingAlgorithmBased2017, hadian-jaziDataReductionSerial2021} has demonstrated proficient peak finding capabilities with minimal user input.  This technique utilizes the Modified Selective Statistical Estimator (MSSE) method to segment background pixels and actual peaks.  Furthermore, the RPF method can also parallelize the processing of diffraction patterns, making it ideal for real-time data analysis.

\psocake{} is a software program specifically designed to streamline high throughput data reduction and analysis of XFEL experiments, and is routinely used at SLAC National Accelerator laboratory (SLAC).  The built-in peak-finding algorithm \citep{shinDataAnalysisUsing2018} includes calibrating raw analog digital units (ADUs); applying optional background correction and bad pixel masks; identifying possible peaks; calculating their SNR; determining Bragg spot sizes; and selecting peaks based on size, total intensity and SNR.

@@

The current peak finding process of \peaknet{} comprises of two steps, as depicted in Fig. \ref{fig : peak finding}.  First, a segmentation map is obtained from the input image using a deep neural network.  Subsequently, peak positions are determined based on the generated segmentation map using connected component analysis \citep{weaverCentrosymmetricCrossSymmetricMatrices1985}.  Importantly, \peaknet{} has the capability to process inputs with arbitrary image sizes in batches, enabling efficient analysis.

In this section, our focus is to provide a comprehensive description of the underlying deep neural network components in \peaknet{}.  We will specifically discuss the network architecture, data augmentation techniques, and the loss function employed during the training process.  Furthermore, we will emphasize the use of iterative data curation in addressing the outlier problem.

\begin{figure*}[!ht]
\includegraphics[width=\textwidth,keepaspectratio]
{./figures/peaknet_steps.pdf}
\caption{The peak finding process of \peaknet{} consists of two steps: (1)
obtaining a segmentation map from the input using the underlying deep neural
network; (2) identifying peak positions based on the generated segmentation map
using connected component analysis.  It is worth noting that \peaknet{} is
capable of accommodating input with arbitrary image sizes and batch sizes.  }
\label{fig : peak finding}
\end{figure*}

@@

Our neural network architecture is primarily based on the original ``Attention U-Net" design \citep{oktayAttentionUNetLearning2018}.  However, we want to point out that the three-component design, consisting of the ``backbone", ``neck" and ``head", offers the flexibility to replace these components with different architectures.  For instance, the ``backbone" can be substituted with a ResNet \citep{heDeepResidualLearning2016}, while the ``neck" can be replaced with other multi-resolution feature fusion methods like feature pyramid networks (FPN) \citep{linFeaturePyramidNetworks2017} or bidirectional feature pyramid networks (BiFPN) \citep{tanEfficientDetScalableEfficient2020}.  Additionally, the ``head" can be extended to perform additional tasks, if desired, such as predicting the likelihood of the input image being indexable.  This modular approach provides flexibility and allows for customization based on specific requirements.

\begin{figure*}[!ht]
\includegraphics[width=\textwidth,keepaspectratio]
{./figures/network_arc.pdf}
\caption{Neural network architecture: Attention U-Net with residual connections.
(a) Overall end-to-end network architecture.  (b) Residual double
convolutional block.  (c) Gated feature fusion block for merging
multi-resolution features.}
\label{fig : network arc}
\end{figure*}

@@

One main technical issue in training neural networks for Bragg peak segmentation is the extreme peak-background class imbalance (e.g., less than $1 : 10^3$), resulting in reduced prediction accuracy.  This may not pose as a problem in previous peak finders like BraggNet, which operates on a much smaller area of $32 \times 32$, \peaknet{} is trained on larger detectors, such as the Rayonix MX340-XFEL detector with a dimension of 1920 $\times$ 1920 pixels even after $4 \times 4$ pixel binning.  To mitigate the label imbalance, we employ a categorical focal loss to address the problem \citep{linFocalLossDense2018}.

\begin{align}
\text{CFL} &= - \sum_{i = 1}^{N}\sum_{j = 1}^{C} 
            \alpha_j \cdot y_j^{(i)} \cdot (1-\hat{p}_j^{(i)})^\gamma 
            \cdot \log{\hat{p}_j^{(i)}}, \\
\hat{p}_j^{(i)} &= f_\theta(x_j^{(i)})
\end{align}

where CFL stands for categorical focal loss, $\alpha_j$ is a balancing factor for each class, $y_j^{(i)}$ is the ground truth label of the $j$-th class for the $i$-th pixel.  $p_j^{(i)}$ is the predicted probability of the $j$-th class for the $i$-th pixel, calculated by the neural network $f$ parameterized by $\theta$ given input image $x_j^{(i)}$ with $N$ number of pixels and $C$ categories.  $\gamma$ is a parameter controlling the extent to which the weight of well-classified examples, such as background pixels, is reduced.  In our neural network training, we chose $\gamma = 2.0$, essentially, this means $(1-\hat{p} _j^{(i)}) ^\gamma$ is solely responsible for down-weighting easy examples, as $\hat{p} _j^{(i)}$ is usually large in this case.  Finding the ideal values for $\alpha$ can be a delicate process with only marginal returns, as highlighted in the original focal loss study \citep{linFocalLossDense2018}.  In order to fully utilize focal loss, we decided to use a fixed trivial value of $\alpha = 1.2$ for all examples, primarily for the sake of completeness.  This choice essentially treats $\alpha$ as a simple scaling factor without significant impact.  In most cases, the sum of all $\alpha_j$ values, denoted as $\sum_{j = 1}^{C} \alpha_j$, sums to 1.

@@

We applied three data augmentation techniques to each diffraction pattern in our dataset, including random in-plane rotation, random shifting in both horizontal and vertical directions and random masking.  Random in-plane rotation and shifting augment the training data, reducing the model's tendency to memorize features at specific locations.  This variability promotes more generalized feature extraction and improved generalization performance.  Random masking involves randomly obscuring parts of the input data, forcing the model to learn other relevant features that enhance its predictive capabilities during training. 50 rectangular masks with random sizes between 80 and 120 were positioned randomly. The result of applying data augmentation to an arbitrary example is shown in Fig. \ref{fig : data aug}.

\begin{figure*}[!ht]
\includegraphics[width=\textwidth,keepaspectratio]
{./figures/data_aug.pdf}
\caption{Illustration of data augmentation on an input image. (a) Augmented
image with random in-plane rotation, shift, and masking. (b) its corresponding
labeled segmentation map. (c, d) Zoomed-in views of (a) and (b), respectively.
The segmentation label distinguishes red pixels as peaks and blue pixels as
artifact scattering, attributed to scattering from metal shielding and injection
nozzle.}
\label{fig : data aug}
\end{figure*}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\columnwidth,keepaspectratio,trim={1in 0in 1in 0in}]
{./figures/data_engine.pdf}
\caption{The three step process of iterative dataset and model improvement. (1) Model learns from training data. (2) Model predictions reveal potential areas of improvement. (3) Correct low-quality training labels or acquire additional examples of challenging instances.}
\label{fig : data engine}
\end{figure}

@@

The primary goal of \peaknet{} is to autonomously detect Bragg peaks in diffraction images, devoid of any user input, such as algorithmic parameter tuning or manual artifact masking.  We implemented a two-step approach for peak detection, employing deep neural networks to perform semantic segmentation, which was subsequently followed by connected component analysis to extract image coordinates of Bragg peaks.  In this section, we begin by introducing the datasets used for training, validating, and testing \peaknet{}. Next, we present the evaluation results that demonstrate the predictive performance of \peaknet{}. Finally, we showcase its auto-masking capability and runtime efficiency.

@@

The dataset ``mfx13016" exhibits a notable ring-like background scattering accompanied by bright Bragg peak-like spots located on the rings.  This characteristic presents two primary challenges for crystallography data analysis: increased difficulty in crystal indexing due to false-positive peaks and inaccurate integration results caused by artifact scattering.  The current best practice involves manual masking of these artifact scattering, enabling data analysis to proceed as if they are absent, but this process is time-consuming and can not adjust to shot-to-shot variations in the artifacts in real-time.  Additionally, the final outcomes often hinge on the expertise of the individuals conducting the data processing, resulting in biases, scalability limitations, lack of standardization, and the possibility of human errors.  Nevertheless, the inherent challenges presented by this dataset offer an exceptional opportunity to showcase the autonomous peak finding capability of \peaknet{}.

@@

The peak finding performance of peak finders is evaluated through three key indicators from a user's standpoint: the number of successfully indexed hits, the indexing rate, and the merging statistics.  For our evaluation, we utilized the CrystFEL software suite \citep{whiteCrystFELSoftwareSuite2012} to perform crystal indexing, integration, and merging tasks.  We use \psocake{}'s peak finder as a baseline in this evaluation.

@@

\peaknet{} and manually fine-tuned \psocake{} exhibit comparable peak finding performance in terms of the number of hits and the number of indexed images, as shown in Table \ref{tb : index}.  However, we emphasize that while \peaknet{} operates entirely autonomously, \psocake{}'s performance is contingent upon manual parameter fine-tuning and manual artifact scattering masking.  For example, an earlier work \citep{suSerialCrystallographyUsing2021}, which processed the same experimental data, reported significantly lower hit count and indexed image count.  Consequently, this highlights \peaknet{}'s advantage in user-independent expert-level data processing.

Moreover, while users might prioritize the absolute number of indexed images as a more straightforward metric, the indexing rate provides valuable insights into the method's efficiency.  Both metrics jointly offer a more comprehensive evaluation of the peak finding method. In the case of \peaknet{}, both consistently high number of indexed images and high indexing rate testify to its effective and reliable autonomous peak finding process.

@@

\multicolumn{1}{l}{\multirow{3}{*}{\peaknet{}}}                                                        & \multicolumn{1}{l}{Number of hits}       & \multicolumn{1}{l}{\textbf{1205}}    & \multicolumn{1}{l}{\textbf{3248}}    & \multicolumn{1}{l}{\textbf{5797}}    & \multicolumn{1}{l}{\textbf{3949}}    & \multicolumn{1}{l}{\textbf{1184}}    & \multicolumn{1}{l}{1030}    & \textbf{1223}    \\ \cline{2-9}
\multicolumn{1}{l}{}                                                                                   & \multicolumn{1}{l}{Number of indexed}    & \multicolumn{1}{l}{\textbf{860}}     & \multicolumn{1}{l}{\textbf{2856}}    & \multicolumn{1}{l}{\textbf{3433}}    & \multicolumn{1}{l}{\textbf{2401}}    & \multicolumn{1}{l}{\textbf{919}}     & \multicolumn{1}{l}{953}     & 1169    \\ \cline{2-9}
\multicolumn{1}{l}{}                                                                                   & \multicolumn{1}{l}{Indexing rate}        & \multicolumn{1}{l}{\textbf{0.71}}    & \multicolumn{1}{l}{\textbf{0.88}}    & \multicolumn{1}{l}{\textbf{0.59}}    & \multicolumn{1}{l}{\textbf{0.61}}    & \multicolumn{1}{l}{0.78}    & \multicolumn{1}{l}{0.93}    & 0.96    \\ \hline

@@

\multicolumn{1}{l}{\multirow{3}{*}{\begin{tabular}[c]{@{}l@{}}\psocake{}\\ (fine-tuned)\end{tabular}}} & \multicolumn{1}{l}{Number of hits}       & \multicolumn{1}{l}{1193}    & \multicolumn{1}{l}{3087}    & \multicolumn{1}{l}{5769}    & \multicolumn{1}{l}{3926}    & \multicolumn{1}{l}{917}     & \multicolumn{1}{l}{\textbf{1070}}    & 1212    \\ \cline{2-9}
\multicolumn{1}{l}{}                                                                                   & \multicolumn{1}{l}{Number of indexed}    & \multicolumn{1}{l}{849}     & \multicolumn{1}{l}{2657}    & \multicolumn{1}{l}{3375}    & \multicolumn{1}{l}{2364}    & \multicolumn{1}{l}{760}     & \multicolumn{1}{l}{\textbf{1005}}    & \textbf{1202}    \\ \cline{2-9}
\multicolumn{1}{l}{}                                                                                   & \multicolumn{1}{l}{Indexing rate}        & \multicolumn{1}{l}{\textbf{0.71}}    & \multicolumn{1}{l}{0.86}    & \multicolumn{1}{l}{\textbf{0.59}}    & \multicolumn{1}{l}{0.60}     & \multicolumn{1}{l}{\textbf{0.83}}    & \multicolumn{1}{l}{\textbf{0.94}}    & \textbf{0.99}    \\ \hline

@@

Merging statistics confirms \peaknet{}'s robust performance in terms of consistency and quality in data processing.  For the thaumatin (TH) sample, \peaknet{} demonstrates marginally superior performance in terms of R-split values, correlation coefficients, $I/\sigma(I)$ values, redundancy, and completeness, achieving these results autonomously.  Notably, the R-split value for thaumatin is slightly lower for \peaknet{} (0.2605 vs. 0.2732), suggesting slightly better internal consistency. Similarly, correlation coefficients for \peaknet{} are marginally higher (CC$_1/2$: 0.9434 vs.  0.9368; CC*: 0.9853 vs. 0.9835), indicating a minor advantage in the precision of measurements.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=\textwidth,keepaspectratio]
{./figures/automask.agbehnate.pdf}
\caption{Examples of identifying scattering rings in silver behenate images using
\peaknet{}.}
\label{fig : automask 2}
\end{figure*}

@@

Peak finding can pose a significant computational demand, particularly in high data rate experiments, highlighting the critical importance of an efficient peak finding algorithm.  A comparative analysis of runtime performance demonstrates the greater efficiency of \peaknet{} compared to other commonly used \psocake{} peak finder.
